Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2020-12-15 manuscript/draft_rsfmri.Rmd
minor changes in manuscript

# [ ] Evaluate SPM model comparison (more activation in SPM..)
# [ ] Discussion
# [ ] hervorheben, was von Präregistrierung abweicht
# [ ] brain images mit nilearn plotting tools einbinden
# [ ] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [ ] re-do analysis with new sample
# [ ] repeat analysis with exclusion of 10% of worst data
# [ ] Network tests

@@

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
```
\newpage
# Abstract

@@

```{r Load data, include=FALSE}
source("code/create_sample_df.R")
final = create_sample_df(group = "all", tp = "all")

@@

# overview distribution of data points
tp1 <- as.character(freqtable$Var1[freqtable$Freq == 1])
tp2 <- as.character(freqtable$Var1[freqtable$Freq == 2])
tp3 <- as.character(freqtable$Var1[freqtable$Freq == 3])
tp1_tab <- table(subset(final, subj.ID %in% tp1)$tp, subset(final, subj.ID %in% tp1)$condition)
tp2_tab <- table(subset(final, subj.ID %in% tp2)$tp, subset(final, subj.ID %in% tp2)$condition)[c(1,3),]
tp3_tab <- table(subset(final, subj.ID %in% tp3)$tp, subset(final, subj.ID %in% tp3)$condition)[1,]
total_datapoints <- c(sum(final$group==1),sum(final$group==2))
total_subjects <- plyr::count(final[subj_idx,"group"])$freq
tab_sample <- data.frame(rbind(tp1_tab, tp2_tab, tp3_tab, total_subjects, total_datapoints))
rm(tp1,tp2,tp3,tp1_tab, tp2_tab, tp3_tab, total_subjects, total_datapoints)
rownames(tab_sample) <- c("count: only 0","count: only 6","count: only 12","count: 0 and 6","count: 6 and 12","count: complete data","total number of subjects","total data points")
colnames(tab_sample) <- c("BARS","NBARS")

@@

```{r tableSample, echo=FALSE}
tab_sample <-
  knitr::kable(
    tab_sample,
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    caption = "Distribution of data points at months after intervention")
tab_sample
```

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/mzyxv](https://osf.io/mzyxv).
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study. <!-- Frage: Ausschluss bei head movement? ADI009_fu2 nicht enthalten (siehe Readme preprocessing) -->
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tableDescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals (n = `r nrow(subj_df)`, `r nrow(filter(subj_df,Sex == "f"))` female; ages `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), that either underwent surgery (n = `r cs["Sum","KG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female). The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.

```{r tableBaselineCharacteristics, include=FALSE}
dfBL <- create_sample_df(group = "all", tp = "BL")

@@

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throuout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tableDescr, echo=FALSE}

# table for descriptive statistics
descr <- psych::describeBy(dfw[,5:ncol(dfw)], group=dfw$condition,mat=T)
descr$predictor <- c(rep("BMI",6),rep("meanFD",6))
descr$tp <- car::recode(descr$vars, "c(1,4)='fu2';c(2,5)='fu';c(3,6)='bl'")
descr$group <- descr$group1
descr <- descr[,c("predictor","tp","group","n","mean","sd")]
descr <- descr[with(descr, order(predictor, tp)),]
descr[,c("mean","sd")] <- round(descr[,c("mean","sd")],2)
descr[c(2:6,8:12),"predictor"] <- ""
descr[c(2,4,6,8,10,12),"tp"] <- ""
rownames(descr) <- NULL

tab_descr <-
  knitr::kable(
    descr,
    row.names = FALSE,
    format = "latex",
    booktabs = T,
    caption = "BMI and mean FD of participants acquired in each condition and time point")

@@

```
<!-- STUDY PROCEDURE -->
Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for STANDARDIZED?? breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
condition.labs <- c("BARS","NBARS")
names(condition.labs) <- c("IG","KG")
# BMI over time each group
figBMIdescr <-
  ggplot(final, aes(
    x = tp,
    y = BMI,
    group = subj.ID,
    color = condition
  )) +
  geom_line() + geom_point() +
  stat_summary(
    aes(group = condition),
    fun = mean,
    geom  = "line",
    size = 2,
    color = c(rep("#00305E", 3), rep("#C00045", 3))
  ) +
  facet_grid(. ~ condition, labeller = labeller(condition=condition.labs)) + xlab("time points")
figBMIdescr +
  scale_x_discrete(labels = c("bl" = "0", "fu" = "6", "fu2" = "12")) +
  scale_color_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")
  ) +
  xlab("Time after intervention [months]") + ylab ("BMI [kg/m²]") +
  theme_bw() +  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "none"
  )
```

@@

#### Functional Quality control
Resting state fMRI data quality control was performed along the protocol of @Ciric_2018.
As the expected average head motion is high in this sample [@Beyer_2020], it is difficult to determine an appropriate exclusion threshold which does not lead to the exclusion of a high number of individuals. Thus, we did not exclude any participants based on mean framewise displacement estimates beyond those who failed the anatomical preprocessing. We applied best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. In addition, for models showing a significant group-by-time interaction, we performed a sensitivity analysis excluding the 10% data sets with highest mean FD (mFD).
On the single-subject level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref{carpet_plot} shows an exemplary carpet plot with (from top to bottom): DVARS (spatial root mean square of the data after temporal differencing), mFD, percent outliers defined by AFNI `OutlierCount`, voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data. All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.
We evaluated mFD-FC relationships by computing a functional connectome based on (Power et al., 2011). In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we correlated mean FD and QC values for each node across participants, and plotted the resulting mFD-FC correlation values in histogram (see Figure \@ref{example_FC}. We reported the mFD-FC correlation per preprocessing scheme.
Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.

@@

Statistical analysis were performed in MATLAB version 9.7.0.1190202 (R2019b) using the development version of the SwE toolbox [@Guillaume_2014] ([https://github.com/NISOx-BDI/SwE-toolbox](https://github.com/NISOx-BDI/SwE-toolbox) as of commit #384d2d7944e0c1919a3d22a0282a513916c25f59 was used because of bugs in most recent release) as implemented in SPM12.7770<!-- spm_check_installation('rev') -->, and R version 3.6.1 [@team2013r].
During analyses an implicit mask and an explicit mask were used. For this, FSL's standard space the brain mask that corresponds th "152 nonlinear 6th generation" atlas was re-sampled to 3 x 3 x 3 mm$^3$.<!-- MNI ICBM 2009a Nonlinear Symmetric grey matter mask ([http://www.bic.mni.mcgill.ca/ServicesAtlases/ICBM152NLin2009](http://www.bic.mni.mcgill.ca/ServicesAtlases/ICBM152NLin2009)) were re-sampled to 3 x 3 x 3 mm$^3$ and binarized at threshold $0.5$. -->

@@

<!-- DESIGN MATRIX, vielleicht beide auch als Abbildung kurz zeigen? --> For the first model, time was introduced as factor because we did not assume strict linearity for the increase in connectivity and therefore refrained from modeling time as continuous variable (see Figure \@ref(fig:figDesignmatrix)). The resulting factorial design <!-- check: flexible factorial? --> contained one regressor for each time point per group.
```{r figDesignmatrix, fig.align = 'center', fig.ncol = 2, out.width = "50%", fig.subcap=c('Design matrix with time as categorial factor.','Design matrix with time as continuous variable.'), fig.cap="Possible design matrices of group-time model.", echo=FALSE}
# fig.show="hold",  out.width="40%", fig.height = 8,

# design matrix time as factor
txt_path <- "/data/pt_02161/Analysis/Project2_resting_state/seed-based/Second_level /SwE_files/noExclFD/both/total/"

ID <- read.delim(paste0(txt_path,"subjID.txt"), head=FALSE)
condition <- read.delim(paste0(txt_path,"group.txt"), head=FALSE)
KG_bl <- read.delim(paste0(txt_path,"KG_bl.txt"), head=FALSE)
KG_fu <- read.delim(paste0(txt_path,"KG_fu.txt"), head=FALSE)
KG_fu2 <- read.delim(paste0(txt_path,"KG_fu2.txt"), head=FALSE)
IG_bl <- read.delim(paste0(txt_path,"IG_bl.txt"), head=FALSE)
IG_fu <- read.delim(paste0(txt_path,"IG_fu.txt"), head=FALSE)
IG_fu2 <- read.delim(paste0(txt_path,"IG_fu2.txt"), head=FALSE)
measurement <- c(1:nrow(ID))
dmf <- data.frame(measurement,ID,condition,KG_bl,KG_fu,KG_fu2,IG_bl,IG_fu,IG_fu2)
colnames(dmf) <- c("measurement","ID","condition","N0","N6","N12","B0","B6","B12")

# design matrix time as continuous variable
ID <- read.delim(paste0(txt_path,"subjID.txt"), head=FALSE)
condition <- read.delim(paste0(txt_path,"group.txt"), head=FALSE)
group_IG <- read.delim(paste0(txt_path,"group_IG.txt"), head=FALSE)
group_KG <- read.delim(paste0(txt_path,"group_KG.txt"), head=FALSE)
tp_IG <- read.delim(paste0(txt_path,"tp_IG.txt"), head=FALSE)
tp_KG <- read.delim(paste0(txt_path,"tp_KG.txt"), head=FALSE)
measurement <- c(1:nrow(ID))
dmc <- data.frame(measurement,ID,condition,group_IG,group_KG,tp_IG,tp_KG)
colnames(dmc) <- c("measurement","ID","condition","groupB", "groupN","timeB","timeN")

# plot dmf
input_dmf <- dmf %>%
  pivot_longer(
    cols= colnames(dmf[,c(4:ncol(dmf))]),
    names_to = "regressor")
input_dmc <- dmc %>%
  pivot_longer(
    cols= colnames(dmc[,c(4:ncol(dmc))]),
    names_to = "regressor")

plotA <- ggplot(input_dmf, aes(x = regressor, y = measurement, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low="black", high="white") +
  labs(x="regressors", y="scans") +
  # design
  scale_x_discrete(expand = c(0,0),labels=colnames(dmf)[4:ncol(dmf)]) +
  scale_y_continuous(expand = c(0,0)) +
  theme_bw() + theme(legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12))
# + theme(panel.grid = element_blank(),
#         panel.border = element_rect(colour = "black", size=1),
#    legend.text = element_text(size = 10)
#         plot.title=element_text(size=11))
plotB <- ggplot(input_dmc, aes(x = regressor, y = measurement, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low="black", high="white") +
  labs(x="regressors", y="scans") +
  # design
  scale_x_discrete(expand = c(0,0),labels=colnames(dmc)[4:ncol(dmc)]) +
  scale_y_continuous(expand = c(0,0)) +
  theme_bw() + theme(legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12))
# + theme(panel.grid = element_blank(),
#         panel.border = element_rect(colour = "black", size=1),
#    legend.text = element_text(size = 10)
#         plot.title=element_text(size=11))

plotA
plotB
#gridExtra::grid.arrange(plotA, plotB, nrow=1)
```
For the second and third model, we calculated the centered mean BMI and log mean FD per subject and the intra-subject-centered BMI as proposed by @Guillaume_2014. With a model containing average BMI and longitudinal change per subject, we investigated their differential effects on the complete sample and subsequently in a exploratory fashion for the intervention group only as the effect of longitudinal change in BMI was expected to be driven by this subsample.
<!-- NUISANCE COVARIATES --> Age at baseline, sex, and log mean framewise displacement (FD) were entered into the first and second model as nuisance variables. Age and sex were between-subject covariates while FD is time-varying. Mean FD was log-transformed and age was demeaned by subtracting the overall mean across subjects and time points.
<!-- Prereg: We will perform sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI -->
As change in FD and change in BMI share common variance [@Beyer_2019], the inclusion of log mean FD may be disputed and done optionally. For that reason, we performed each analysis separately one with mean FD (three and without. Differences in results will be reported. We consider this sensitivity analysis theoretically more meaningful than the preregistered comparison with the nuisance covariate free model and a model with four nuisance covariates (including baseline BMI).
<!-- POST HOC  --> For a better understanding of the unique contribution of average and longitudinal change FD measures, we employed a third post-hoc model to disentangle the effects: $FC = between-subject FD + within-subject FD$. These post-hoc analyses deviate from the preregistration and should be considered to be exploratory.
<!-- INFERENCE --> <!-- Prereg: We will use Type 3 small samples bias adjustment and “Approx” option for estimating the degrees of freedom. We will use the Non-Parametric Wild Bootstrap approach with default parameters, 5000 bootstraps and clusterwise inference (cluster forming threshold of p<0.01). Significant clusters are defined as family-wise error - corrected p<0.05. -->
To ensure robustness of our results, we computed estimates with non-parametric test using Wild bootstrap with an unrestricted SwE on all contrasts of interest for clusterwise inference. Type C2 for small sample bias adjustment is applied before the bootstrap resampling as recommended in the SwE manual. In contrast to the parametric estimation, Wild Bootstrap avoids any parametric distributional assumptions (e.g. random field theory), and is unlike permutation based framework appropriate<!-- permutations don't work with multilevel data (dependencies)--> for longitudinal data.
Unlike the description in the preregistration, we used a cluster forming threshold of $p < 0.001$ for a more multiple comparison adjustment (instead of $p < 0.01)$, and $1000$ bootstraps due to required computation time (instead of 5000). Significant clusters are defined as family-wise error (FWE) corrected $p < 0.05$.

<!-- R analysis of aggregated values -->
We analyzed the aggregate connectivity values using linear mixed models in the R package lme4.
Following the preregistration, we first investigated a time-by-group interaction between baseline and followup.
We used a model comparison of R1 and R0 models with `anova`:
`R1= lmer(agg FC ~ timepoint*group + (1|subj))`
`R0 = lmer(agg FC ~ timepoint + group + (1|subj))`
We performed a sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI. Again, the group-by-time point interaction was considered significant if a model comparison between R1 and Null models using anova-function yielded a p<0.05.
Finally, we estimated the above-mentioned interaction model for all three time points. Again, the interaction was considered significant if a model comparison between R1 and R0 using `anova` had p<0.05.

@@

## Quality control
We performed different quality controls for the rsfMRI data.
As described in @Beyer_2020, there was a significant interaction of group and time on mFD. In particular, the intervention group reduced in head motion in the follow-up assessments compared to the control group.
We also inspected DVARS, a measurement of whole-brain variation in signal which is often reflective of noise in the data. We saw that DVARS did not qualitatively differ between groups and timepoints, and that the correlation of DVARS and mFD remained similar over time points.
This indicates that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period.
```{r tableQA, fig.cap="Summary table of QA measures"}
for (i in 1:nrow(final)) {
  tmp = strsplit(toString(final[i, "scan_dir"]), '/')[[1]][8]
  conf = read.csv(
    paste0("/data/pt_02161/preprocessed/resting/detailedQA/metrics/",tmp,"/confounds.csv")
  )
  tmp = cor.test(conf$DVARS, conf$FD)
  final[i, "corr_DVARS_FD"] = tmp$estimate
  final[i, "mean_DVARS"] = mean(conf$DVARS, na.rm = TRUE)
  final[i, "max_DVARS"] = max(conf$DVARS, na.rm = TRUE)
  final[i, "mean_outliers"] = mean(conf$outliers, na.rm = TRUE)
}

>>>>

# [ ] Evaluate SPM model comparison (more activation in SPM..)
# [ ] Discussion
# [ ] brain images mit nilearn plotting tools einbinden
# [ ] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [x] re-do analysis with new sample
# [ ] repeat analysis with exclusion of 10% of worst data
# [ ] Network tests

@@

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex") 
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
```
\newpage
# Abstract

@@

```{r "Load data", include=FALSE}
source("report/create_sample_df.R")
final = create_sample_df(group = "all", tp = "all")

@@

@@

```{r "tableSample", echo=FALSE}
source("./report/tables.R")
tab_sample <- mk_SampleTable(final)

tab_sample <-
  knitr::kable(
    tab_sample,
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "Distribution of data points at months after intervention")
tab_sample
```

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/mzyxv](https://osf.io/mzyxv).
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study.
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tableDescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals (n = `r nrow(subj_df)`, `r nrow(filter(subj_df,Sex == "f"))` female; ages `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), that either underwent surgery (n = `r cs["Sum","KG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female). 
The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r tableBaselineCharacteristics, include=FALSE}
dfBL <- create_sample_df(group = "all", tp = "BL")

@@

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tableDescr, echo=FALSE}

# table_descr <- mk_tableDescr()
# table for descriptive statistics
descr <- psych::describeBy(dfw[,5:ncol(dfw)], group=dfw$condition,mat=T)
descr$predictor <- c(rep("BMI",6),rep("meanFD",6))
descr$tp <- car::recode(descr$vars, "c(1,4)='fu2';c(2,5)='fu';c(3,6)='bl'")
descr$group <- descr$group1
descr <- descr[,c("predictor","tp","group","n","mean","sd")]
descr <- descr[with(descr, order(predictor, tp)),]
descr[,c("mean","sd")] <- round(descr[,c("mean","sd")],2)
descr[c(2:6,8:12),"predictor"] <- ""
descr[c(2,4,6,8,10,12),"tp"] <- ""
rownames(descr) <- NULL

tab_descr <-
  knitr::kable(
    descr,
    row.names = FALSE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "BMI and mean FD of participants acquired in each condition and time point")

@@

```
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).
```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
source("report/figures.R")
figBMIdescr <- mk_figBMIdescr(final)

@@

#### Functional Quality control
Resting state fMRI data quality control was performed along the protocol of @Ciric_2018.
As the expected average head motion is high in this sample [@Beyer_2020], it is difficult to determine an appropriate exclusion threshold which does not lead to the exclusion of a high number of individuals. Thus, we did not exclude any participants based on mean framewise displacement estimates beyond those who failed the anatomical preprocessing. We applied best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitored the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. In addition, for models showing a significant time-by-group interaction, we performed a sensitivity analysis excluding the 10% data sets with highest mean FD (mFD).
On the single-subject level, we used `plot_carpet` from `niworkflows.viz.plots` to generate carpet plots which depict denoising quality. Figure \@ref{carpet_plot} shows an exemplary carpet plot with (from top to bottom): DVARS (spatial root mean square of the data after temporal differencing), mFD, percent outliers defined by AFNI `OutlierCount`, voxelwise timeseries from minimally preprocessed, AROMA non-aggressive and aggressive denoising, CC and CC + GSR data. All carpet plots were visually checked and if there was still structured noise in the CC+GSR-denoised functional data (such as black/white stripes, signal dropout), the participant was given a rating of 1 in `QA residuals`.
We evaluated mFD-FC relationships by computing a functional connectome based on [@Power_2012]. In more detail, we used the MNI coordinates of 246 cortical and subcortical spherical seed regions with 5mm radius in `NiftiSpheresMasker`, and calculated the Spearman rank correlation between these time series. Then, we correlated mean FD and QC values for each node across participants, and plotted the resulting mFD-FC correlation values in histogram (see Figure \@ref{example_FC}. We reported the mFD-FC correlation per preprocessing scheme.
Further, we evaluated the distance-dependence of mFD-FC correlations by correlating the euclidean distance between nodes with the mFD-QC correlation and report mean, median and standard deviations of mFD, DVARS and number of outliers over all participants.

@@

Statistical analysis were performed in MATLAB version 9.7.0.1190202 (R2019b, @MATLAB2018) using the development version of the SwE toolbox [@Guillaume_2014] ([https://github.com/NISOx-BDI/SwE-toolbox](https://github.com/NISOx-BDI/SwE-toolbox) as of commit #384d2d7944e0c1919a3d22a0282a513916c25f59 was used because of bugs in most recent release) as implemented in the Statistical Parametric Mapping software (SPM12.7770, @spm12)<!-- spm_check_installation('rev') -->, and R version 3.6.1 [@team2013r].
During analyses an implicit mask and an explicit mask were used. For this, FSL's standard space the brain mask that corresponds th "152 nonlinear 6th generation" atlas was re-sampled to 3 x 3 x 3 mm$^3$.<!-- MNI ICBM 2009a Nonlinear Symmetric grey matter mask ([http://www.bic.mni.mcgill.ca/ServicesAtlases/ICBM152NLin2009](http://www.bic.mni.mcgill.ca/ServicesAtlases/ICBM152NLin2009)) were re-sampled to 3 x 3 x 3 mm$^3$ and binarized at threshold $0.5$. -->

@@

<!-- DESIGN MATRIX, vielleicht beide auch als Abbildung kurz zeigen? --> For the first model, time was introduced as factor because we did not assume strict linearity for the increase in connectivity and therefore refrained from modeling time as continuous variable (see Figure \@ref(fig:figDesignmatrix)). The resulting factorial design <!-- check: flexible factorial? --> contained one regressor for each time point per group.
```{r figDesignmatrix, fig.align = 'center', fig.ncol = 2, out.width = "50%", fig.subcap=c('Design matrix with time as categorial factor.','Design matrix with time as continuous variable.'), fig.cap="Possible design matrices of time-by-group model.", echo=FALSE}
# fig.show="hold",  out.width="40%", fig.height = 8,
source("report/figures.R")
PlotList <- mk_figDesignMatrix()

PlotList$plotA
PlotList$plotB
#gridExtra::grid.arrange(plotA, plotB, nrow=1)
```
For the second and third model, we calculated the centered mean BMI and log mean FD per subject and the intra-subject-centered BMI as proposed by @Guillaume_2014. With a model containing average BMI and longitudinal change per subject, we investigated their differential effects on the complete sample and subsequently in a exploratory fashion for the intervention group only as the effect of longitudinal change in BMI was expected to be driven by this subsample.
<!-- NUISANCE COVARIATES --> Baseline age, sex, and log mean framewise displacement (FD) were entered into the first and second model as nuisance variables. Age and sex were between-subject covariates while FD is time-varying. Mean FD was log-transformed and age was demeaned by subtracting the overall mean across subjects and time points. 
As change in FD and change in BMI share common variance [@Beyer_2019], the inclusion of log mean FD may be disputed and done optionally. For that reason, we performed each analysis separately one with mean FD (three and without. Differences in results will be reported. We consider this sensitivity analysis theoretically more meaningful than the preregistered comparison with the nuisance covariate free model and a model with four nuisance covariates (including baseline BMI).<!-- Prereg: We will perform sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI -->
<!-- POST HOC  --> For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc model to disentangle the effects: $FC = between-subject FD + within-subject FD$ with nuisance covariates age and sex. As these post-hoc analyses were not preregistered, they should be considered to be exploratory.

<!-- INFERENCE --> To ensure robustness of our results, we computed estimates with non-parametric test using Wild bootstrap with an unrestricted SwE on all contrasts of interest for clusterwise inference. 
Type C2 for small sample bias adjustment is applied before the bootstrap resampling as recommended in the SwE manual. 
In contrast to the parametric estimation, Wild Bootstrap avoids any parametric distributional assumptions (e.g. random field theory), and is unlike permutation based framework appropriate<!-- permutations don't work with multilevel data (dependencies)--> for longitudinal data.
Unlike the description in the preregistration, we used a cluster forming threshold of $p < 0.001$ for a more multiple comparison adjustment (instead of $p < 0.01)$, and $1000$ bootstraps due to required computation time (instead of 5000). Significant clusters are defined as family-wise error (FWE) corrected $p < 0.05$. The anatomical localization of significant clusters was investigated with the SPM Anatomy toolbox, version 2.2c [@Eickhoff_2005].<!-- Prereg: We will use Type 3 small samples bias adjustment and “Approx” option for estimating the degrees of freedom. We will use the Non-Parametric Wild Bootstrap approach with default parameters, 5000 bootstraps and clusterwise inference (cluster forming threshold of p<0.01). Significant clusters are defined as family-wise error - corrected p<0.05. -->

<!-- R analysis of aggregated values -->
We analyzed the aggregate connectivity values using linear mixed models in the R package lme4 [@Bates_2007].
Following the preregistration, we first investigated a time-by-group interaction between baseline and follow-up.
We used a model comparison of R1 and R0 models with `anova`:
`R1= lmer(agg FC ~ timepoint*group + (1|subj))`
`R0 = lmer(agg FC ~ timepoint + group + (1|subj))`
We performed a sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI. Again, the time-by-group point interaction was considered significant if a model comparison between R1 and Null models using anova-function yielded a p<0.05.
Finally, we estimated the above-mentioned interaction model for all three time points. Again, the interaction was considered significant if a model comparison between R1 and R0 using `anova` had p<0.05.

@@

## Quality control
We performed different quality controls for the rsfMRI data.
As described in @Beyer_2020, there was a significant interaction of group and time on mFD. In particular, the intervention group reduced in head motion in the follow-up assessments compared to the control group.
We also inspected DVARS, a measurement of whole-brain variation in signal which is often reflective of noise in the data. We saw that DVARS did not qualitatively differ between groups and time points, and that the correlation of DVARS and mFD remained similar over time points.
This indicates that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period.
```{r tableQA, fig.cap="Summary table of QA measures"}

for (i in 1:nrow(final)) {
  tmp = strsplit(toString(final[i, "scan_dir"]), '/')[[1]][8]
  conf = read.csv(
    paste0("/data/pt_02161/preprocessed/resting/detailedQA/metrics/",
      tmp, "/confounds.csv"))
  tmp = cor.test(conf$DVARS, conf$FD)
  final[i, "corr_DVARS_FD"] = tmp$estimate
  final[i, "mean_DVARS"] = mean(conf$DVARS, na.rm = TRUE)
  final[i, "max_DVARS"] = max(conf$DVARS, na.rm = TRUE)
  final
}
